The Evolution of Phenotypically Invariant Genetic Networks

### 2020-06-01 review-responses.tex

Your manuscript is quite long already, and the reviewers' suggestions of
modifications and clarification, which need to be implemented, will make the
manuscript even longer. I however share the reviewers' opinion that the
manuscript's two parts are only loosely related. My suggestion is therefore to
publish them separately. Although connecting the systems biology and the
quantitative genetics parts is indeed an exciting endeavor, this is not really
achieved in the current version of the manuscript, and it would be more
profitable to first better describe each part separately. Regarding the
connexion between the two parts, please also pay particular attention to R2's
first specific comment about dimensionality.

This may be because I am more familiar with Fisher's geometric model than
systems biology models, but it seems to me that the first part is more novel,
and should therefore deserve your attention first, should you follow my
suggestion of publishing the two parts separately.
\end{quote}

\begin{point}{Format:}
    Please add line numbers to your manuscript; it is straightforward
with the lineno package in LaTeX.  Please also do not increase text width too
much, as this decreases legibility (LaTeX's default settings already optimize
the number of characters per line).  Finally, be careful about not confusing
citep and citet (a lot of missing parentheses around citations in the second part).
\end{point}

\reply{
  We've added line numbers and corrected improperly formatted citations. 
}

@@

\begin{point}{all figures:}
    Please ensure that all axes legends and labels are big enough
and not distorted (same of arrow labels on the diagrams). Please also make sure
that the lines are thick enough to be visible.
\end{point}

\reply{
}

\begin{point}{Above eq (2):}
    ``Under quite general assumptions'': quickly mention them?
\end{point}

\reply{
  Thank you for the suggestion. We now mention some \revref.
  \jss{I think the general assumptions are time invariance and linearity? Any others we should mention?}
}

@@

\begin{point}{p5, Note on implementation}
    ``the closure of span(B)'' may not be
    understandable by most of Evolution's readers; please explain what this means.
\end{point}

\reply{
    We've added a clarifying note \revref
    but we're hoping that readers not familiar with linear algebra will realize
    that they can skip this ``note on implementation'' without loss of continuity.
}

\begin{point}{Figures 1-3:}
    In figures 1 and 3, there are two types of arrow heads (+ and -
effects), but not in Figure 2, probably because the sign of the effect depends
on the value of $\tau$. Maybe use a  different type of arrow heads for
interactions that may change signs, to avoid the confusion with positive
effects?
\end{point}

\reply{
  We think that adding a third type of arrowhead may only add to the confusion, so now we are using only the standard arrowhead \revref.
  \jss{we could add a cirlce arrowhead for the unknown direction, but I think its simpler to just get rid of the rectangular negative arrowhead originally in Figure 3}
}

@@

\begin{quote}
General remark: line numbers are very helpful for reviewing.  
\end{quote}

\begin{quote}
    \textbf{(I)} The first (systems biology) part makes use of the concept of
``nonidentifiability'' of chemical reaction networks and develops these concepts
in the context of a genotype to phenotype map. In particular, this map is
analyzed for the case where it can be modeled as a linear system.
\end{quote}

\begin{point}{}
    First note that the concept of ``nonidentifiability'' is closely related to
themes that have been discussed in evolutionary biology for a long time, under
names like ``mutational robustness'', ``network neutrality'', ``canalization'',
``redundancy'' etc. In particular, if you characterize your model by ``many
distinct (and mutationally connected) molecular pathways can realize identical
phenotypes'', note that this essentially describes what has been called a
``neutral network'' and studied under this name in many articles. Also concrete
developmental networks have been studied in this context (eg, von Dassow etal
2000, Nature 406:188–192). Discussion of this literature - and more importantly
discussion of terms and notions used here relative to terms that have been used
elsewhere - is largely missing. I won't be able to summarize all this here, but
an older review is the Evolution Perspective piece ``Evolution and detection of
genetic robustness'' (de Visser et al 2003). You can work work backward and
forward from there.  
\end{point}

\reply{
  We've reworked the introduction and included a paragraph on the phenonmenon of biological degeneracy, which unites many of these concepts. \revref
}

@@

\begin{point}{}
    Related to this: I find it non-intuitive to start the introduction with the
notion of ``nonidentifiability'', which is a consequence rather than a cause, the
cause being what you later call ``phenotypic equivalence'' and what had been
discussed under other names in other articles.
\end{point}

\reply{
  Good point. In the introdcution we now introduce nonidentifiability after discussing general cases of phenotypic equivalence. \revref
}

@@

\begin{point}{}
    Still related: You give long lists of references in the introduction, but do
not provide the reader with any information about specific contributions. A bit
more would be helpful (starting with True and Haag coined the term
``developmental system drift'', etc).
\end{point}

\reply{
  We've reworked the introduction and included more detail about these references. \revref
}

@@

\reply{
}

\begin{point}{}
    Population isolates and genetic load: Isn't this exactly ``founder effect
speciation''?
\end{point}

\reply{
}

\begin{point}{}
    Your analysis of the fitness loss in F1 and F2 hybrids rests on a local
Taylor expansion of the fitness landscape. This is adequate given that the
underlying linear network is also only locally valid (see above). However, you
then apply this to a discussion of hybrid incompatibility (Haldane's rule etc).
We are thus interested in \emph{very large} fitness costs for hybrids. This does not
seems to be compatible: to get hybrid incompatibility or sterility, you need to
have epsilon sufficiently large (below equation 5). But then $\epsilon^2$ is no
longer smaller than $\epsilon$. In other words: it seems to me that for the
discussion of hybrid incompatibilities you apply your model of the first part
beyond the local range where it is valid.
\end{point}

\reply{
    That's a good point; we resort to a Taylor expansion 
    because we couldn't think of any other clear way to get quantitative estimates.
    However, our simulations XXX.
}

### 2020-06-01 kalman_walk_paper.tex
first pass 2020

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm, xfrac}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}
\usepackage{multicol}
\usepackage[bottom]{footmisc}

@@

\begin{document}
%\linenumbers

@@

It is an overarching goal of many biological subdisciplines 
to attain a general understanding of the function and evolution of the 
complex molecular machinery that translates an organism's genome 
into the characteristics on which natural selection acts,
the phenotype.
% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
%Attaining a general understanding of the functioning and evolution of this molecular machinery
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.
%The dynamics of the molecular machinery and its interactions with the environment
%can be mathematically described as a dynamical system \citep{jaeger2015comet}.
%Movement in this direction is ongoing, as 
There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
and empirically inspired
\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
% If we allow the reasonable assumption that the genotype-phenotype map 
% can be represented as a system of differential equations \citep{jaeger2015comet}, 
% we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

It is well known that in many contexts mathematical models
can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modelled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
%even if a one-to-one mapping between model and behavior is possible in theory,
%even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.

Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
but viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.

The main purpose of this paper is to explore this general idea
in the concrete framework of linear systems theory,
thought of as modeling gene regulatory networks.
First, we apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
This suggests that, quite generally, there are many directions in which a gene network can drift
without changing the phenotype.
Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
system drift may result in reproductive incompatibility between isolated populations,
even in the absence of any sort of adaptive, or environmental change.

 % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

% wright1932roles mentions possibility of equivalent combinations
% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
% barton1989divergence following on from barton1986maintenance analyzing multiple optima in a similar model, looking at RI due to shifts between equilibria
% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%% Fisher's geom model: (incompat due to segregation variance)
% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
%       hybrids will suffer because of cryptic segregating variation
% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
%        but mostly only if fitness is exp(-|x|^k) for k > 2
% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
%   and (b) mutational effects fall under the CLT
% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
%%%% General notes
% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

@@

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2.2$} (B);
   % \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
    \path [->, sloped] (A) edge[bend right] node {\tiny $2$} (C); 
    %\path [->, sloped] (B) edge[bend right] node {\tiny ?} (C); 
    \path [->, >=Rectangle, sloped] (C) edge[bend right] node {\tiny $-1$} (A); 
    \path [->, >=Rectangle, sloped] (C) edge[bend right] node {\tiny $-2.2$} (B); 
    \path[->] (U) edge node {\tiny $1$} (A);
    \path[->] (U) edge node {\tiny $1$} (B);
    \path[->] (A) edge node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop below] node[sloped, below] {\tiny $1$} (A);
    \path [->] (B) edge[loop above] node[sloped, above] {\tiny $1$} (B);
    \path [->, >=Rectangle] (C) edge[loop right] node[sloped, above] {\tiny $-1$} (C);
\end{scope}
\end{tikzpicture}
  \caption{A possible non-minimal three-gene oscillator, phenotypically equivalent to $A(\tau)$, the systems in Examples \ref{ex:oscillator} and \ref{ex:all_osc}.}
  \label{fig:3D_osc}
\end{center}
\end{figure}

@@

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs to be occupied for transcription to begin),
and variation in expression level measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless, these measures likely give us an idea of the scale of variability.
%\citet{kasowski2010variation} found 
It has been shown that between human individuals, there is differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) \cite{kasowski2010variation}.
%\citet{verlaan2009targeted} showed that 
It has also been inferred that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells \cite{verlaan2009targeted}, and
%while \citet{lappalainen2013transcriptome} found 
that human population variation 
explained about 3\% of expression variation \cite{lappalainen2013transcriptome}.
Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
%\citet{wang2017bayesian} found 
allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species has been observed \cite{wang2017bayesian}, as well as
%while \citet{tung2015genetic} observed 
allele-specific expression in 23.4\% of genes studied in a baboon species \cite{tung2015genetic}. 
Taken together, this suggests that variation in the entries of $A$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm, xfrac}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, allcolors=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}
\usepackage{multicol}
\usepackage[bottom]{footmisc}

@@

\begin{document}
\linenumbers

@@

 Biological systems, at nearly all levels of organization, display considerable \emph{degeneracy} -- 
that is systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}.
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick, woesse}
all the way up to the convergent evolution of adaptive traits. Between these levels, degeneracy can be found in the \emph{neutral networks}
  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner, andmore}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide},
as well as in the thermodynamic stability of proteins \citep{hart2014thermodynamic}. Further examples include the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
and neural circuits \citep{trojanowski}, and in developmental systems \citep{true2001developmental}. 
\revpoint{1}{1} \revpoint{1}{3}

Likewise, it is well known that in many contexts mathematical models
can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modeled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes
referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
  tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}. \revpoint{1}{2}

In this paper we use linear systems theory to model gene regulatory networks, linking biological degeneracy 
with mathematical nonidentifiability. Although our focus is on linear systems, nonidentifability is by no 
means unique to such frameworks, and is a general property of many nonlinear models \citep{nonlinear}.

System degeneracy has been argued to be necessary for natural selection \citep{edleman2001degeneracy}, 
as it may function as a mechanism for biological robustness and evolvability \citep{robustness}, or manifest as 
\emph{canalization} \citep{canalization}. Degeneracy of the genetic code, and of genetic networks, 
for instance, can make sequences more fault-tolerant to mutations and open up pathways to adaptation (cite). 
Biological degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura},
and perhaps even contribute to the formation of new species \citep{gavrilets}.

We are interested in the nature and evolutionary implications of degeneracy and nonidentifiability in the context of gene regulatory networks.
Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If systems architectures are not functionally unique, can this open up neutral evolutionary paths, and do
  these paths manifest as \emph{developmental system drift} \citep{true2001developmental} and contribute to speciation?
First, we apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
Finally, we briefly outline and apply a quantitative genetics framework to estimate the speed at which system drift can lead to speciation via reproductive incompatibility,
which we develop in more detail in a companion paper \citep{schiffman2020genetic}.

%It is an overarching goal of many biological subdisciplines 
%to attain a general understanding of the function and evolution of the 
%complex molecular machinery that translates an organism's genome 
%into the characteristics on which natural selection acts,
%the phenotype.
%For example, 
%there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
%\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
%as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
%Mathematical models of both particular regulatory networks
%and the evolution of such systems in general
%can provide guidance where intuition fails,
%and thus has the potential to discover general principles in the organization of biological systems 
%as well as provide concrete numerical predictions \citep{servedio2014not}.
%There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
%both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
%and empirically inspired
%\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
%
%It is well known that in many contexts mathematical models
%can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
%there can be uncertainty about an inferred model's parameters or even its claims about
%causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
%Models with different parameter schemes, or even different mechanics 
%can make equally accurate predictions,
%but still not actually reflect the internal dynamics of the system being modelled.
%In control theory, where electrical circuits and mechanical systems are often the focus, 
%it is understood that there can be an infinite number of ``realizations'', 
%or ways to reverse engineer the dynamics of a ``black box'',
%even if all possible input and output experiments are performed 
%\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
%The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
%In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
%Finally,
%the field of \emph{inverse problems} studies those cases in which,
%despite the existence of a theoretical one-to-one mapping between a model and behavior,
%tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
%
%Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
%but viewed from another angle,
%this concept can provide a starting point for thinking about externally equivalent systems
%-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
%These functional symmetries manifest in convergent and parallel evolution, 
%as well as \emph{developmental system drift}: the observation that
%macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
%\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
%
%The main purpose of this paper is to explore this general idea
%in the concrete framework of linear systems theory,
%thought of as modeling gene regulatory networks.
%First, we apply results from system theory which give
%an analytical description of the set of 
%all linear gene network architectures that yield identical phenotypes.
%This suggests that, quite generally, there are many directions in which a gene network can drift
%without changing the phenotype.
%Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
%system drift may result in reproductive incompatibility between isolated populations,
%even in the absence of any sort of adaptive, or environmental change.
%We then briefly outline a quantitative genetics framework,
%which we further develop in a companion paper,
%which estimates how quickly system drift (and more generally, the presence of a fitness ridge)
%leads to reproductive incompatibility \citep{schiffman2020genetic}.
%
% % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
%% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%%
%%AND THESE?
%  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
%  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.  
%
%% wright1932roles mentions possibility of equivalent combinations
%% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
%% barton1989divergence following on from barton1986maintenance analyzing multiple optima in a similar model, looking at RI due to shifts between equilibria
%% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
%% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
%% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
%% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
%% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
%% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
%% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%%% Fisher's geom model: (incompat due to segregation variance)
%% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
%% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
%%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
%% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
%%       hybrids will suffer because of cryptic segregating variation
%% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
%%        but mostly only if fitness is exp(-|x|^k) for k > 2
%% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
%%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
%%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
%% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
%% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
%%   and (b) mutational effects fall under the CLT
%% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
%%%%% General notes
%% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic
%
%It is not a new observation that there is often more than one way to do the same thing, 
%or that speciation can be the result of (nearly) neutral processes.
%The potential for speciation has been analyzed in models of 
%traits under stabilizing selection determined additively by alleles at many loci 
%\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
%in related fitness landscape models \citep{fraisse2016genetics},
%and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
%% Recent work has shown that Fisher's geometric model of stabilizing selection
%% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
%It has also been shown that population structure
%can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
%However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
%and a regulatory cascade \citep{porter2002speciation}
%found rapid speciation under directional selection
%but only equivocal support for speciation under models of purely neutral drift.
%%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
%The rate at which hybrid incompatibility accumulates due to genetic drift creating
%segregation variance between isolated populations is fairly well understood
%\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
%but model assumptions can strongly affect predictions,
%including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
%and the shape of the fitness landscape \citep{fraisse2016genetics}.
%Our main aim is to provide a concrete framework 
%that can provide natural predictions of these model parameters
%across a general class of models.
%Furthermore, tools from system theory allow analytical predictions to be made
%for large populations with complex phenotypes
%that would be inaccessible to population simulations.
%

@@

\revpoint{AE}{7}

@@

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2.2$} (B);
   % \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
    \path [->, sloped] (A) edge[bend right] node {\tiny $2$} (C); 
    %\path [->, sloped] (B) edge[bend right] node {\tiny ?} (C); 
    \path [->, sloped] (C) edge[bend right] node {\tiny $-1$} (A); 
    \path [->, sloped] (C) edge[bend right] node {\tiny $-2.2$} (B); 
    \path[->] (U) edge node {\tiny $1$} (A);
    \path[->] (U) edge node {\tiny $1$} (B);
    \path[->] (A) edge node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop below] node[sloped, below] {\tiny $1$} (A);
    \path [->] (B) edge[loop above] node[sloped, above] {\tiny $1$} (B);
    \path [->] (C) edge[loop right] node[sloped, above] {\tiny $-1$} (C);
\end{scope}
\end{tikzpicture}
  \caption{A possible non-minimal three-gene oscillator, phenotypically equivalent to $A(\tau)$, the systems in Examples \ref{ex:oscillator} and \ref{ex:all_osc}.}
  \label{fig:3D_osc}
\end{center}
\end{figure}

@@

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs to be occupied for transcription to begin),
and variation in expression level measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless, these measures likely give us an idea of the scale of variability.
%\citet{kasowski2010variation} found 
It has been shown that between human individuals, there is differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) \citep{kasowski2010variation}.
%\citet{verlaan2009targeted} showed that 
It has also been inferred that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells \citep{verlaan2009targeted}, and
%while \citet{lappalainen2013transcriptome} found 
that human population variation 
explained about 3\% of expression variation \citep{lappalainen2013transcriptome}.
Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
%\citet{wang2017bayesian} found 
allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species has been observed \citep{wang2017bayesian}, as well as
%while \citet{tung2015genetic} observed 
allele-specific expression in 23.4\% of genes studied in a baboon species \citep{tung2015genetic}. 
Taken together, this suggests that variation in the entries of $A$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.
